% arXiv-ready LaTeX version of docs/paper.html (pdflatex-compatible).
% Thai script is transliterated; the rendered-Thai version is docs/paper.pdf in the repo.
\documentclass[10pt]{article}
\usepackage[a4paper,margin=22mm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}

\title{\bfseries Cost-Aware Evaluation of Multi-Agent LLM Systems\\for Thai Sarcasm Detection}
\author{Thanaphumi Kunuthai\\
\small\url{https://github.com/Thanaphumi2006/Cost-Aware-Multi-Agent-LLM-Evaluation-for-Thai-Sarcasm-Detection}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Multi-agent LLM architectures (verifier pipelines, debate panels, hybrid committees) are widely
assumed to justify their extra inference cost with higher quality. We test that assumption on a task
where surface form and intent diverge sharply: sarcasm detection in Thai (\emph{prachot}).
On a 127-item human-labeled gold set drawn from product reviews and tweets, we compare five systems
under identical measurement code, tracking F1, API calls, dollar cost, and latency simultaneously.
A two-agent screener--verifier pipeline initially appears to win (F1 0.744 vs.\ 0.690). The advantage
does not survive a fair baseline: letting the single agent read its own token-level confidence and
tuning one threshold, chosen leave-one-fold-out at zero added cost, lifts it to 0.725, after which
no multi-agent system separates from it (best $\Delta$F1 $+0.019$, 95\% CI $[-0.073, +0.113]$,
paired bootstrap). Model choice dominates architecture choice: a model $6\times$ cheaper than the
flagship stays within a 0.05 F1 band, making a cheap single agent statistically indistinguishable
from a flagship pipeline at one-eleventh the cost. We further establish a \$0 floor: three
independent free methods (a 3-character regex, an embedding kNN, and a fine-tuned Thai transformer)
all converge at F1 $\approx$ 0.59, and show a three-way confidence router that retains 95\% of
flagship quality at 40\% of the cost. In the deployed cascade the pattern recurs once more: a free
abstention threshold on the lexical tier (F1 $0.627 \to 0.806$ at 42\% of the calls) beats inserting
a trained model tier, which a constant-output control ties exactly. Expanding the gold set $4.7\times$ (595 items, 104 positive)
with adversarially plausible negatives narrows the pipeline-vs-tuned-baseline confidence interval
from 0.186 to 0.056 wide and preserves the null result, with the point estimate now favouring the
single agent ($-0.019$, 91\% chance the pipeline is not better) at $2.96\times$ the cost. At that
scale the confidence-reading upgrade itself becomes statistically significant for the first time
($+0.040$, 95\% CI $[+0.018, +0.060]$), while precision collapses for every architecture equally:
hard negatives are a property of the data that additional agents do not repair. On a uniformly random test set ($4.9\%$ sarcastic, the true base rate
versus 17--24\% in the mined sets), every headline number collapses: the free cue detector falls
from F1 0.59 to 0.09 and argmax GPT flags 36\% of all text (F1 0.20), while the free
confidence-reading upgrade is what rescues the system to F1 0.46, its largest single gain anywhere.
We conclude that for subjective, low-resource classification, the first questions are not
architectural: exhaust the information one agent already returns, then choose the cheapest sufficient
model.
\end{abstract}

\section{Introduction}
Multi-agent LLM systems promise quality through redundancy: multiple calls, multiple roles, a final
adjudication. The promise is easy to believe and expensive to hold; every added agent multiplies API
cost and latency. Whether the quality materializes is an empirical question, and one that is rarely
asked with the cost axis in view.

We ask it on Thai sarcasm detection. Sarcasm (\emph{prachot}) is a deliberately adversarial target:
the writer states the opposite of their intent, so surface cues mislead exactly the models that rely
on them. Thai adds a low-resource setting, informal orthography (vowel elongations, the laughter
numeral ``555''), and, in our source domains of restaurant reviews and tweets, a culture of
exuberant sincere praise that is easy to mistake for irony. If multi-agent deliberation helps
anywhere, a task defined by hidden intent should be favorable ground.

Our contributions:
(1) a human-labeled Thai sarcasm gold set with a sharp operational rubric (sarcasm requires
\emph{feigned praise}), documented provenance and bias notes, plus a $4.7\times$-expanded hard
variant labeled blind through a purpose-built tool;
(2) a five-system, four-axis comparison (F1, calls, cost, latency) under one measurement harness,
with paired bootstrap and McNemar tests throughout;
(3) a negative result with an identified mechanism: multi-agent gains over a
\emph{confidence-aware} single agent are statistically unresolvable while costing 1.8--7.4$\times$,
because half of the apparent gain was the baseline discarding confidence it had already paid for;
(4) cost-axis anchors: a \$0 floor at F1 $\approx$ 0.59 reached independently by three free methods,
and a confidence router giving 95\% of flagship quality at 40\% of its cost.

\section{Task and Data}
\textbf{Definition.} An item is sarcastic (label 1) iff the author feigns praise or thanks while
intending the opposite. The rubric resolves the three boundary cases that dominate disagreement:
direct complaints are 0 (negative content without pretense), balanced reviews are 0 (genuine praise
plus genuine criticism), and exuberant sincere praise is 0. Undecidable items (context-dependent
intent) are excluded rather than guessed.

\textbf{Gold set.} 72{,}865 texts from Wongnai restaurant reviews and the Wisesight social-media
corpus were keyword-scored for sarcasm suspicion; 400 sampled items were pre-labeled by GPT-4o and
then human-reviewed \emph{blind} (the annotator decides before seeing the model's draft). Agreement
analysis showed the LLM over-flags sarcasm: of 17 LLM-positive items in the first review pass, the
human kept 3, motivating blind review as a design rule. Iterative mining of high-suspicion texts
brought the canonical set to \textbf{127 items (30 sarcastic / 97 not)}. Provenance caveats (some
positives were mined with the same model family later evaluated, likely inflating its recall) are
documented and affect all systems equally.

\textbf{Hard expansion.} A later blind-labeling campaign exhausted the mined candidate pool
(495 items), giving \textbf{gold\_v2: 595 items (104 sarcastic, 17.5\%)}. Because candidates were
pre-selected to \emph{look} sarcastic and 81\% were judged sincere, gold\_v2 is deliberately
enriched with plausible negatives: a hard set whose absolute scores are not comparable to the
canonical set (\S\ref{sec:repl}).

\section{Systems}
All LLM systems use the same detector prompt and JSON-constrained output on GPT-4o unless stated;
prices are July-2026 list prices.
\textbf{S1 Single agent}: one call per item, argmax label.
\textbf{S1$+\tau$ Threshold}: identical calls, but the positive-class probability is read from token
logprobs and thresholded; $\tau$ is chosen leave-one-fold-out (stratified 5-fold, seed 42), so no
item influences the threshold that judges it.
\textbf{S2 Pipeline}: a screener flags candidates; a \emph{constrained} verifier may only overturn
$1\to0$ and only when confident, preserving the screener's recall.
\textbf{S3 Debate}: prosecutor and defender argue; a judge decides freely (3 calls/item).
\textbf{S4 Hybrid}: a four-agent committee mixing the above roles.
\textbf{S5 WangchanBERTa}: a Thai transformer fine-tuned on the gold set under 5-fold
cross-validation $\times$ 3 seeds (every prediction from a model that never saw the item), run
locally at \$0.

\section{Evaluation Protocol}
Accuracy is not reported: with 76\% negatives, constant rejection scores 0.76. All systems predict
the same items, so comparisons are paired: F1 differences carry 95\% CIs from a paired bootstrap
(5{,}000 resamples of the item set), and label flips are counted with McNemar's test. Costs are
computed from measured token usage; latency is wall-clock per item (p50).

\section{Results}

\begin{table}[h]\centering\small
\begin{tabular}{lcccccc}
\toprule
System & F1 & Prec. & Rec. & Calls & Cost & p50 \\
\midrule
S1 Single agent (argmax) & 0.690 & 0.526 & \textbf{1.000} & 127 & \$0.094 & 751 ms \\
S1$+\tau$ + threshold & 0.725 & 0.641 & 0.833 & 127 & \textbf{\$0.094} & 751 ms \\
S2 Two-agent pipeline & \textbf{0.744} & 0.604 & 0.967 & 183 & \$0.169 & 967 ms \\
S3 Debate (3 agents) & 0.694 & 0.595 & 0.833 & 381 & \$0.695 & 4{,}557 ms \\
S4 Hybrid (4 agents) & 0.700 & 0.560 & 0.933 & 292 & \$0.407 & 832 ms \\
S5 WangchanBERTa (local) & 0.620 & 0.553 & 0.700 & \textbf{0} & \textbf{\$0.00} & \textbf{26 ms} \\
\bottomrule
\end{tabular}
\caption{All systems on the same 127 items, same measurement code.}
\end{table}

\subsection{The fair baseline erases the multi-agent win}
Against the plain single agent, the pipeline's $+0.054$ looks meaningful (95\% CI $[-0.002,
+0.113]$, P(not better) = 3\%). But that baseline throws away the model's own confidence. Reading it
back, at zero extra calls and zero extra dollars, lifts the single agent to 0.725, and against
\emph{that} baseline every multi-agent system is unresolved: pipeline $+0.019$ $[-0.073, +0.113]$ at
$1.8\times$ cost; hybrid $-0.025$ $[-0.117, +0.071]$ at $4.3\times$; debate $-0.030$ $[-0.158,
+0.096]$ at $7.4\times$. Roughly half of the apparent multi-agent gain was information the baseline
had already paid for and discarded.

\subsection{Architecture: constraints beat head-count}
Why does the two-agent pipeline outperform the three-agent debate ($\Delta$ $+0.049$, at a quarter
of debate's price)? The detector starts from recall 1.000; what remains is buying precision without
spending recall. The pipeline's verifier is \emph{structurally constrained}: it can only overturn
positive verdicts, and only when explicitly confident, so it keeps recall at 0.967 while pruning
false positives. Debate returns full freedom to the panel: its defender successfully argues away 5
true positives (recall 0.833), spending the asset the detector had secured. The number of agents
matters less than the constraint structure imposed on them.

\subsection{Model choice dominates architecture choice}
The tuned single agent across five OpenAI models spanning a $25\times$ price range stays within a
0.05 F1 band, with the best score from one of the cheapest (gpt-4.1-mini: F1 0.727, \$0.015). That
cheap single agent is statistically tied with the flagship pipeline (gpt-4o, 0.744, \$0.169) at
\textbf{1/11 of the cost}. Local open models (Qwen2.5-7B, SeaLLM-7B, Typhoon-8B, OpenThaiGPT-7B,
4-bit, chat-templated, same leave-fold-out protocol) top out at 0.576: a real but modest gap of
$+0.152$ to the GPT bot, and Thai-specialized models show no advantage over general Qwen.

\subsection{Anchoring the cost axis}
\textbf{The \$0 floor.} Three independent free methods (the regex ``555'' alone, F1 0.590; an e5
embedding kNN, 0.588; fine-tuned WangchanBERTa at the default threshold, 0.590) converge at
$\approx 0.59$. The floor is a property of the task, not of any one method.
\textbf{A useful middle.} A three-way router (confident-negative / uncertain $\to$ escalate to GPT /
confident-positive) at a 40\% escalation budget reaches F1 0.690, 95\% of the GPT bot, for 40\% of
its cost; beyond that the curve is flat. A cautionary side-result: tuning WangchanBERTa's threshold
leave-fold-out \emph{underperforms} not tuning at all (0.556 vs.\ 0.590) at this $n$; at small
sample sizes, a tunable knob is a liability, and the 0.071 ``gain'' from tuning on the full set is
leakage.

\subsection{The deployed cascade: a free abstention knob beats a model tier}
The shipped tool escalates only cue-uncertain items to the LLM. Does inserting the fine-tuned
WangchanBERTa as a free middle tier (cue $\to$ WangchanBERTa $\to$ GPT) improve it? Measured offline
on the 127-item set with leakage-free out-of-fold WangchanBERTa probabilities and cached gpt-4.1-mini
scores, the outcome splits into a real lever and an inert one. The lever is a free \emph{abstention}
threshold on the cue tier: letting the lexical model answer only when its combined log-lift exceeds
$\log 2.46$ (one strong cue) and escalating the rest lifts F1 from 0.627 to \textbf{0.806} while
paying for 42\% of the calls, above GPT-on-everything (0.746) at under half its cost and stable under
cross-validation ($0.802 \pm 0.023$). The inert component is the model tier itself: replacing every
WangchanBERTa decision with the constant ``not sarcastic'' under identical routing yields an
\emph{identical} 0.806; the middle tier's paired $\Delta$F1 over cue$\to$GPT is 0.000; its answers
are not seed-stable (0.769 / 0.806 / 0.806 across three seeds); and the confident-negative threshold
calibrated out-of-fold does not transfer. On ten unseen sentences the deployed model's sarcasm
probability never falls below 0.37, so the tier fires on none of them. This is a third instance of
the paper's pattern: a free knob that gates the confidence already present outperforms inserting an
additional model.

\subsection{Replication at 4.7$\times$ the data}\label{sec:repl}

\begin{table}[h]\centering\small
\begin{tabular}{lcccc}
\toprule
System (gold\_v2, $n=595$) & F1 & Prec. & Rec. & Cost \\
\midrule
S1 Single agent (argmax) & 0.337 & 0.206 & 0.933 & \$0.275 \\
\textbf{S1$+\tau$ + threshold (LOFO)} & \textbf{0.377} & 0.238 & 0.894 & \textbf{\$0.275} \\
S2 Two-agent pipeline & 0.357 & 0.222 & 0.913 & \$0.814 ($2.96\times$) \\
\bottomrule
\end{tabular}
\caption{Core systems on the expanded hard set. Absolute scores are not comparable to Table~1 by
design.}
\end{table}

On the expanded set two things resolve. First, the confidence-reading upgrade that defines our fair
baseline is now \emph{statistically significant}: $+0.040$ F1 over the plain baseline, 95\% CI
$[+0.018, +0.060]$, McNemar 80:6, at identical cost and call count. What was a suggestive trend at
$n=127$ and $n=302$ is confirmed at $n=595$. Second, the thesis comparison holds and hardens:
measured against that baseline the pipeline is $\mathbf{-0.019}$ (95\% CI $[-0.047, +0.009]$), a
91\% chance it is not better, McNemar 49:83 in the single agent's favour, at $2.96\times$ the cost.
The interval narrows monotonically with sample size (0.186 at $n=127$, 0.094 at $n=302$, 0.056 at
$n=595$) without the sign of the conclusion ever flipping toward multi-agent.

Absolute F1 falls to ${\approx}0.35$ for every system because the added items are adversarially
plausible negatives: 403 of the 495 mined candidates (81\%) were judged sincere by the human
annotator, diluting the positive rate to 17.5\% and surging false positives identically across
architectures. The model is also more overconfident at this scale (404/595 items outside
$[0.01, 0.99]$), yet the threshold still extracts a significant gain, and its leave-fold-out score
(0.377) sits within 0.004 of the oracle-tuned value (0.381), indicating a stable rather than
overfitted cutoff.

\subsection{Reality check on a uniformly random test set}
Every set above was keyword-mined to 17--24\% sarcastic. To measure the true base rate we drew 250
uniformly random texts from the 72{,}865-item pool, labeled blind: only \textbf{4.9\%} (12/245
decidable) are sarcastic, so keyword mining inflates the positive rate roughly $4.5\times$. On this
realistic distribution the class-imbalance mathematics dominates every result.

\begin{table}[h]\centering\small
\begin{tabular}{lcccc}
\toprule
System (random set, $n=245$, 4.9\% sarc.) & F1 & Prec. & Rec. & Flag rate \\
\midrule
Always ``not sarcastic'' & 0.000 & --- & 0.000 & 0\% \\
Lexical cues (deployed free tool) & 0.090 & 0.055 & 0.250 & 22\% \\
GPT-4o (argmax) & 0.202 & 0.115 & 0.833 & 36\% \\
\textbf{GPT-4o + threshold (LOFO)} & \textbf{0.462} & 0.429 & 0.500 & 6\% \\
\bottomrule
\end{tabular}
\caption{Absolute performance on a realistically distributed test set. Threshold F1 95\% CI
$[0.182, 0.688]$ (wide: only 12 positives).}
\end{table}

Two anchors from the mined data are exposed as artifacts. The ``\$0 floor at F1 0.59'' (the free cue
detector) collapses to \textbf{0.09}: the same cues that fire on sarcasm in a 24\%-positive set fire
overwhelmingly on sincere text when positives are rare (52 false positives), so precision falls to
0.055. This is not a property of the method but the arithmetic of a rare positive class. GPT with
argmax fails identically, flagging 36\% of all text as sarcastic (precision 0.115) when only 4.9\% is.

The one intervention that survives is the free confidence-reading upgrade that defines our fair
baseline. Thresholding the logprob cuts the flag rate from 36\% to 6\% (near the true 4.9\% base
rate), lifting precision from 0.115 to 0.429 and F1 from 0.202 to \textbf{0.462}: a $+0.26$ gain at
zero added cost, larger than any architectural change measured in this work. On realistic data,
reading the confidence already paid for is not a marginal upgrade but the line between a usable and
an unusable system. The best absolute F1 achievable here is ${\sim}0.46$, well below the 0.73
headline from the mined set; sarcasm detection on real traffic is materially harder than any mined
benchmark implies.

\section{Limitations}
(1) \textbf{Scale.} 30 positives (canonical) cannot resolve ${\sim}0.02$ F1 differences; the
expansion raises power (interval width 0.186 to 0.056) but is not i.i.d.
(2) \textbf{Provenance bias.} Some positives were mined with the model family under evaluation; the
bias is shared across systems, so paired comparisons stand, but absolute recall should be read
cautiously.
(3) \textbf{Domain transfer.} On 55 hand-labeled Pantip forum comments, precision falls
$0.68 \to 0.40$ (sincere praise flagged as sarcasm) while recall holds; deployment outside
reviews/tweets requires re-tuning on target-domain labels.
(4) \textbf{Single language/task.} Conclusions are demonstrated for Thai sarcasm; the \emph{method}
(fair confidence-aware baselines, paired statistics, cost as a first-class axis) is the
transferable part.

\section{Conclusion}
On a task chosen to favor deliberation, multi-agent LLM systems did not separate from a single agent
that simply reads its own confidence, while costing 1.8--7.4$\times$ more. The result replicated at
double the data with threefold-tighter intervals. Before adding agents, exhaust the one you have:
read its confidence, tune with leakage-free protocol, and choose the cheapest sufficient model. The
same economy governs a deployed cascade's cheap tiers: a free abstention threshold on the lexical
screener beat inserting a trained model between it and the LLM, which a constant-output control
matched exactly. Architecture is a question to ask after the information already purchased has been
spent.

\begin{thebibliography}{9}\small
\bibitem{du2023} Du, Y., Li, S., Torralba, A., Tenenbaum, J.~B., and Mordatch, I. (2023).
Improving Factuality and Reasoning in Language Models through Multiagent Debate.
\emph{arXiv:2305.14325}.
\bibitem{efron1993} Efron, B. and Tibshirani, R.~J. (1993). \emph{An Introduction to the
Bootstrap}. Chapman \& Hall.
\bibitem{joshi2017} Joshi, A., Bhattacharyya, P., and Carman, M.~J. (2017). Automatic Sarcasm
Detection: A Survey. \emph{ACM Computing Surveys}, 50(5).
\bibitem{lowphansirikul2021} Lowphansirikul, L., Polpanumas, C., Jantrakulchai, N., and Nutanong,
S. (2021). WangchanBERTa: Pretraining transformer-based Thai language models.
\emph{arXiv:2101.09635}.
\bibitem{mcnemar1947} McNemar, Q. (1947). Note on the sampling error of the difference between
correlated proportions or percentages. \emph{Psychometrika}, 12(2).
\bibitem{pythainlp2019} PyThaiNLP (2019). Wisesight Sentiment Corpus.
\emph{github.com/PyThaiNLP/wisesight-sentiment}.
\bibitem{wongnai2018} Wongnai (2018). Wongnai Review Dataset.
\emph{huggingface.co/datasets/wongnai/wongnai\_reviews}.
\end{thebibliography}

\end{document}
